Why does preregistration increase the persuasiveness of evidence? A Bayesian rationalization.

### 2022-02-19 README.Rmd
save

---
output: github_document
---

@@

# Why does preregistration increase the persuasiveness of evidence? A Bayesian rationalization.

Preregistration is becoming increasingly popular in the psychological sciences.
At the same time, many researchers struggle to make precise predictions and feel overwhelmed by the need to prespecify every detail of their analysis plan.
Should they surrender and discard the idea of preregistration altogether? Not at all.
We argue for the benefits of preregistration beyond strictly confirmatory studies.
Grounded in Bayesian philosophy of science, we define a formal objective for preregistration that neither declares changes to a preregistration to be sinful nor punishes rigor.
Our argument rests on the relevance of “theoretical risk”, a generalization of type-I error rate, for judging the evidential support for theories.
Our proposal is that preregistration should aim to reduce the uncertainty in judging the theoretical risk instead of directly limiting the type-I error rate.
We argue that this reformulation extends the utility of preregistration and is closer aligned with researchers intuition.
If we do not know what a researcher did, we have great uncertainty about the theoretical risk they took.
This uncertainty severely limits the utility of the researchers' evidence.

A successful prediction should lend more credibility to a theory than a postdiction, all else equal.
Much has been theorized on this important distinction between exploration and confirmation, discovery and justification, hypothesis generating and hypothesis testing and so forth (see XXXX).
To honor this distinction, many psychological studies are preregistered to produce results that are considered confirmatory.
Rigorous application of preregistration prevents researcher from reporting a simple confirmatory story for a set of results that were produced by an arduous process of try and error.
The scientific community, widely agrees that actively concealing such "trying for results" is misconduct.
But how is such concealment hurting the scientific process?
We believe that a rigorous answer to this question enables us to clear some common misunderstandings that lead researchers to deem preregistration as unfit for their purposes.
These misunderstandings are interesting from a theoretical perspective but also lead to decreased efficiency of the scientific process if researcher misguidedly decide that preregistration is not applicable for their intended study.

@@

Deciding if a study is confirmatory or exploratory is a judgment made by the researcher.
The aforementioned issues arise by delegating this judgment to a simple decision rule, that is, if something is preregistered it is confirmatory and if it is not preregistered it is exploratory.
This rule is often assumed implicitly as a rule of thumb, other times it is stated as the explicit goal of preregistration, but always leads to confusion when researchers try to apply it outside specific bounds.
Preregistering an analysis that is inherently exploratory (like testing dozens of relations) does not make it confirmatory nor will a careful confirmatory study become exploratory if researcher deviate from the preregistration in minor details.
To uphold the simplilicity of the decision rule, it is routinely suggested to only preregister when a study is expected to have a low type I error rate by design and will run/be analysed without changes no matter what.
Under these conditions the rule holds, but such restricted use makes preregistration a niche solution not able to match the greater problem of replicability in psychology and elsewhere.

We show that this simple decision rule is a special case of a more general conceptualization under Bayesian reasoning.
To that end, we first introduce some basic tools of Bayesian philosophy of science and map the exploration/confirmation distinction onto a dimensional quantity we call "theorethical risk", that is inversely related to type I error rate.
We than propose two interpretations of how theorethical risk is influenced by preregistration.
The first interpretation corresponds to the traditional application of preregistration to research paradigms that focus on confirmation by maximizing the theoretical risk or equivalently limiting type I error.
The second interpretation, is our main contribution, and shows broad applicability of preregistration for both exploratory and confirmatory studies that are implemented as preregistered or have undergone changes.
Under this interpretation theoretical risk is not necessarily directly maximized by preregistration but the uncertainty in judging the theoretical risk is minimized.
This changed viewpoint is crucial for the following reason.
Directly maximizing theoretical risk does, as the name implies, increase the risk that a study does fail to deliver results in favor of the theory.
Minimizing the uncertainty around the theoretical risk, as we will later show, shares the beneficial properties but without increasing likelihood of a "failed" study.
Let us note, that a study that is not able to find evidence that a researcher is hoping for, is still immensely valuable for the scientific process.
Under the current incentive structure, however, many researchers have reason to avoid such outcome and few, if any, have the privalige to operate outside of these incentives.
We therefore propose an interpretation of preregistration that increases the expected epistemic value of studies without necessarily increasing the theoretical risk.

## The epistemic value of studies from a bayesian perspective

@@

The first assumption leads to our adoption of a Bayesian framework.
Our rational is as follows.
Researchers that decide to conduct a study are akin to choosing a study to bet on.
They have to "place the bet" by conducting the study therefore invest resources and stand to gain epistemic value with some probability.
This conceptualization of choosing a study to conduct as a betting problem, lets us apply a "Dutch Book" argument.
This argument, states that any better must follow the axioms of probability if they want to avoid being "irrational", i.e., accept bets with sure losses.
Fully developing a Dutch book argument for this problem requires a careful consideration of what kind of studies to include as possible bets, defining a conversion rate from the stakes to the reward, and modeling what liberties researchers have in choosing a study.
Without deliberating these concepts further, we find it persuasive that researchers should not violate the axioms of probability if they have some expectation about what they stand to gain with some likelihood from conducting a study.
The axioms of probability are all that is required to derive the famous Bayes formula, on which we will heavily rely for our arguments.
Please note that our decision to adopt this aspect of Bayesian philosophy of science does not imply anything about the statistical methods researcher use.
In fact, this conceptualization is purposefully reductionistic to be compatible with a wide range of philosophies of science and statistical methods researchers might subscribe to.
Our conclusions can be applied to bayesian and frequentist methods alike.
Of course, frequentist methods are traditionally viewed from a poperian philosophy of science, but our conclusions are compatible with this view and and notable later advancements like the error statistical view.
However, those philosophies of science require a much richer set of assumptions that are not quite necessary for the points we want to make about the value of preregistration.

## Relating Riskiness to epistemic value

@@

Let us make some simplifying assumptions and define notation.
We restrict ourselfs to consider evidence of a binary nature (either exists or does not), since continues evidence would led to some quite involved derivations.
We denote the probability of a theory before observing evidence as $P(H)$ and its complement as $P(¬H) = 1 - P(H)$.
The probability of observing evidence under some hypothesis is $P(E|H)$.
We can calculate the probability of the hypothesis after observing the evidence with help from the Bayes formula:

@@

There we can replace the total probability of evidence, $P(E)$ with $P(H)P(E|H) + P(¬H)P(E|¬H)$ and arrive at:

$$
P(H|E) = \frac{P(H)P(E|H)}{P(H)P(E|H) + P(¬H)P(E|¬H)}
$$

@@

From the three probabilities involved in the calculation, $P(E|¬H)$ is the one our argument rests on.
It tells use how likely it is that we find evidence, when in fact the theory is not true.
How $P(E|¬H)$ relates to $P(H|E)$ is of great relevance, since $P(H|E)$ is often used directly or indirectly as a measure of corroboration of a hypothesis.
In carnaps tradition, in its direct use, it is called corroboration as firmness, in its relation to $P(H)$ it is called increase in firmness.
We refrain from discussing specific measures of corroboration, since there is no measure that show universally better properties than others.
However, it is generally expected that any measure of corroboration increases monotonically with an increase in $P(H|E)$.

In short, we want to increase $P(H|E)$.
Increases in $P(H|E)$ are associated with increases epistemic value, of which we want to maximize the expectation.
So how can we increase $P(H|E)$?
The Bayes formula gives us three options to investigate, namely $P(H)$, $P(E|H)$ and $P(E|¬H)$.
Before we continue, let us note that $P(E|H)$ and $P(E|¬H)$ are not complements (they do not necessarily sum to 1).
The first option leads us to the unsurprising conclusion that higher $P(H)$ leads to higher $P(H|E)$, but the a priori probability of a hypothesis is nothing our study design can change.
The second option is similar commonsense, that is an increase $P(E|H)$ leads to higher $P(H|E)$.
Researcher therefore should make sure that they design their studies in a way that in case their hypothesis is true, they can actually find evidence for it.
While this truism is strongly related to study design, it is unrelated to preregistration.
The third option, while mathematically equally straightforward, is not quite as intuitive.
It concerns $P(E|¬H)$, therefore the probability of observing evidence for the theory, when in fact the theory is wrong.
Before we discuss what this probability means and how it relates to preregistration, let us show how it impacts $P(H|E)$.
First, increasing $P(E|¬H)$ does decrease $P(H|E)$.

## How statistical method contribute to $P(E|¬H)$

For the overly simplistic case that the hypothesis is a statistical nill-hypothesis, $P(E|¬H)$ becomes the type I error rate.
Researchers who choose a smaller type I error rate, can therefore be surer of their results.
However, the research hypothesis is seldomly equal to the statistical null hypothesis.
$P(E|¬H)$ encompasses factors also outside of the statistical realm, like the studies design.
Statistical methods have the great advantage that we have a large toolbox of assessing and controlling their contribution to $P(E|¬H)$.
Examples of our ability to exert control are the setting of type I error rate, the use of corrected fit measures (i.e., adjusted R²) or cross validation in machine learning.
These are all tools that help us to factor in the tendency of statistical methods to show results even when in fact their are none and are therefore closely related to the probability of observing evidence when in fact there is none.

@@

Observe, that since $P(E|¬H)$

>>>>

---
title: "Why does preregistration increase the persuasiveness of evidence? A Bayesian rationalization."
output:
  bookdown::pdf_document2:
    latex_engine: "xelatex"
    toc: false
linestretch: 2
abstract: |
  Preregistration is becoming increasingly popular in the psychological sciences.
  At the same time, many researchers struggle to make precise predictions and feel   overwhelmed by the need to prespecify every detail of their analysis plan.
  Should they surrender and discard the idea of preregistration altogether?
  Not at all.
  We argue for the benefits of preregistration beyond strictly confirmatory studies.
  Grounded in Bayesian reasoning, we define a formal objective for   preregistration that neither declares changes to a preregistration to be sinful   nor punishes rigor.
  This objective rests on the relevance of “theoretical risk”, a generalization of   type-I error rate, for judging the evidential support for theories.
  Our proposal is that preregistration should aim to reduce the uncertainty in   judging the theoretical risk instead of directly limiting the type-I error rate.
  We argue that this reformulation extends the utility of preregistration and is   closer aligned with researchers intuition.
  If we do not know what a researcher did, we have great uncertainty about the   theoretical risk they took.
  This uncertainty severely limits the utility of the researchers' evidence.
---

@@

A successful prediction should lend more credibility to a theory than a postdiction, all else equal.
Much has been theorized on this important distinction between exploration and confirmation, discovery and justification, hypothesis generating and hypothesis testing and so forth (see XXXX).
To honor this distinction, many psychological studies are preregistered to produce results that are considered confirmatory.
Rigorous application of preregistration prevents researcher from reporting a simple confirmatory story for a set of results that were produced by an arduous process of try and error.
The scientific community, widely agrees that actively concealing such "trying for results" is misconduct --- but how exactly is it impeding scientific progress?
We believe that a rigorous answer to this question enables us to clear some common misunderstandings that lead researchers to deem preregistration as unfit for their purposes.
We believe that a rigorous answer to this question enables us to clear some common misunderstandings that lead researchers to deem preregistration as unfit for their purposes.
These misunderstandings are interesting from a theoretical perspective but also lead to decreased efficiency of the scientific process if researcher misguidedly decide that preregistration is not applicable for their intended study.

@@

Deciding if a study is confirmatory or exploratory is a judgment made by the researcher assessing it.
The aforementioned issues arise by delegating this judgment to a simple decision rule, that is, if something is preregistered it is confirmatory and if it is not preregistered it is exploratory.
This rule is often assumed implicitly as a rule of thumb, other times it is stated as the explicit goal of preregistration.
In both cases it leads to confusion when applied outside specific bounds.
Preregistering an analysis that is inherently exploratory (like testing dozens of relations) does not make it confirmatory nor will a carefully conducted confirmatory study become exploratory if researcher deviate from the preregistration in minor details.
To uphold the simplicity of the decision rule, it is routinely suggested to only preregister when a study is expected to have a low type I error rate by design and will run/be analysed without changes no matter what.
Under these conditions the rule holds, but such restricted use makes preregistration a niche solution not able to match the greater problem of replicability in psychology and elsewhere.

We show that this simple decision rule is a special case of a more general conceptualization under Bayesian reasoning.
To that end, we first introduce some basic tools of Bayesian philosophy of science and map the exploration/confirmation distinction onto a dimensional quantity we call "theorethical risk", that is inversely related to type I error rate.
We than propose two interpretations of how theorethical risk is influenced by preregistration.
The first interpretation corresponds to the traditional application of preregistration to research paradigms that focus on confirmation by maximizing the theoretical risk or equivalently limiting type I error.
The second interpretation, is our main contribution, and shows broad applicability of preregistration for both exploratory and confirmatory studies that are implemented as preregistered or have undergone changes.
Under this interpretation theoretical risk is not necessarily directly maximized by preregistration but the uncertainty in judging the theoretical risk is minimized.
This changed viewpoint is crucial for the following reason.
Directly maximizing theoretical risk does, as the name implies, increase the risk that a study does fail to deliver results in favor of the theory.
Minimizing the uncertainty around the theoretical risk, as we will later show, shares the beneficial properties but without increasing likelihood of a "failed" study.
Let us note, that a study that is not able to find evidence that a researcher is hoping for, is still immensely valuable for the scientific process.
Under the current incentive structure, however, many researchers have reason to avoid such outcome and few, if any, have the privilege to operate outside of these incentives.
We therefore propose an interpretation of preregistration that increases the expected epistemic value of studies without necessarily increasing the theoretical risk.

# The epistemic value of studies from a bayesian perspective

@@

The first assumption leads to our adoption of a Bayesian framework.
Our rational is as follows.
Researchers that decide to conduct a study are akin to choosing a study to bet on.
They have to "place the bet" by conducting the study therefore invest resources and stand to gain epistemic value with some probability.
This conceptualization of choosing a study to conduct as a betting problem, lets us apply a "Dutch Book" argument.
This argument, states that any better must follow the axioms of probability if they want to avoid being "irrational", i.e., by accepting bets which lead to sure losses.
Fully developing a Dutch book argument for this problem requires a careful consideration of what kind of studies to include as possible bets, defining a conversion rate from the stakes to the reward, and modeling what liberties researchers have in choosing a study.
Without deliberating these concepts further, we find it persuasive that researchers should not violate the axioms of probability if they have some expectation about what they stand to gain with some likelihood from conducting a study.
The axioms of probability are all that is required to derive the famous Bayes formula, on which we will heavily rely for our arguments.
Please note that our decision to adopt this aspect of Bayesian philosophy of science does not imply anything about the statistical methods researcher use.
In fact, this conceptualization is purposefully reductionistic to be compatible with a wide range of philosophies of science and statistical methods researchers might subscribe to.
Our conclusions can be applied to bayesian and frequentist methods alike.
Of course, frequentist methods are traditionally viewed from a poperian philosophy of science, but our conclusions are compatible with this view and and notable later advancements like the error statistical view.
However, those philosophies of science require a much richer set of assumptions that are not quite necessary for the points we want to make about the value of preregistration.

# Theorethical Risk and Epistemic Value

@@

Let us make some simplifying assumptions and define notation.
We restrict ourselfs to consider evidence of a binary nature (either exists or does not), since continues evidence would led to some quite involved derivations.
We denote the probability of a hypothesis before observing evidence as $P(H)$ and its complement as $P(¬H) = 1 - P(H)$.
The probability of observing evidence under some hypothesis is $P(E|H)$.
We can calculate the probability of the hypothesis after observing the evidence with help from the Bayes formula:

@@

$P(H|E)$ is of great relevance, since $P(H|E)$ is often used directly or indirectly as a measure of corroboration of a hypothesis.
In carnaps tradition, in its direct use, it is called corroboration as firmness, in its relation to $P(H)$ it is called increase in firmness.
We refrain from discussing specific measures of corroboration, since there is no measure shows universally better properties than others.
However, it is generally expected that any measure of corroboration increases monotonically with an increase in $P(H|E)$.

In short, we want to increase $P(H|E)$.
Increases in $P(H|E)$ are associated with increased epistemic value, of which we want to maximize the expectation.
So how can we increase $P(H|E)$?
The Bayes formula gives us three options to investigate, namely $P(H)$, $P(E|H)$ and $P(E)$.
The first option leads us to the unsurprising conclusion that higher $P(H)$ leads to higher $P(H|E)$, but the a priori probability of a hypothesis is nothing our study design can change.
The second option is similar commonsense, that is an increase $P(E|H)$ leads to higher $P(H|E)$.
<!--this sentence is shit-->
Researcher therefore should make sure that they design their studies in a way that in case their hypothesis is true, they can actually find evidence for it.
While this truism is strongly related to study design, it is unrelated to preregistration.
So $P(E)$ remains to consider.
Since $P(E)$ is the denominator, increasing it will decrease $P(H|E)$.
So the unlikelier it is to observe evidence, the more it increases the probability of the hypothesis if we do observe it.
In other words high risk, high reword.

If we equate riskiness with low probability of evidence, the bayesian rational is perfectly in line with the observation that risky prediction do lead to persuasive evidence.
This tension between high risk leading to high reward is central to our consideration of preregistration.
A high risk, high reword strategy, is bound to result in many losses that are eventually absorbed by the high gains.
Sustaining many "failed" studies is not exactly aligned with the incentive structure many researchers operate under.
Therefore researchers have an incentive to appear to take more risks than they actually do, which misleads their colleges to give their claims more credence than they deserve.
It is here, where preregistrations make their impact.
They try to ensure a proper judgment of the riskyness of a study, by making tempering with the appearance of $P(E)$ difficult.

To gain a better understanding on how preregistrations can do that, let us take a closer look at what contributes to $P(E)$.
Using the law of total probability we can split $P(E)$ in two terms:

$$
P(E) = P(H)P(E|H) + P(¬H)P(E|¬H)
$$

We already have noted that there is not much to do about $P(H)$ (and hence its counterprobability $P(¬H)$) and that is common sense to increase $P(E|H)$.
The real lever to pull is therefore $P(E|¬H)$.
This probability tells use how likely it is that we find evidence, when in fact the theory is not true.
Its counterprobability $P(¬E|¬H)= 1 - P(E|¬H)$ is what we call "theorethical risk", because it is the risk a theory takes on in making a prediction about the occurrence of particular evidence in its favor.

@@

Let us note some interesting properties about theoretical risk $P(¬H|¬E)$.
First, increasing theorethical risk leads to higher $P(H|E)$ (our objective).
Second, if the theoretical risk is smaller than $P(E|H)$ it follows that $P(H|E)$ must decrees when observing the evidence.
Third, if the theorethical risk equals zero, than $P(H|E)$ can at best be equal to $P(H)$ but only if $P(H|E)$ is one.
In other words observing a sure fact does not lead credence to a hypothesis.

This sounds like a truism but is directly related to Poppers famous criterion of demarcation.
He stated, that if it is impossible to proof a hypothesis false, therefore $P(E|¬H) = 1$ or conversely $P(¬E|¬H) = 0$, it can not be a scientific hypothesis (Popper 1962, 39).
We note these relations to underline that the Bayesian rational we apply here is able to reconstruct many commonly hold views on riskiness and epistemic value.

# Should Preregistration Increase the Theorethical Risk?

Now after we discussed that increasing the theoretical risk will increase the epistemic value it is only natural to task preregistration with maximizing theoretical risk.

# How Statistical Methods Influence the Theorethical Risk

Preregistration is often discussed in its relation to statistical methods, so let us note some relations between $P(E|¬H)$ and statistical methods.
If you consider the overly simplistic case where the research hypothesis is equal to the statistical nill-hypothesis, than the alternative Hypothesis is $¬H$, and therefore $P(E|¬H)$ is equivalent to the type I error rate.
Researchers who choose a smaller type I error rate, can therefore be surer of their results.
However, the research hypothesis is seldomly equal to the statistical null hypothesis.
We argue that $P(E|¬H)$ also encompasses factors outside of the statistical realm most notably the study design and broader analytical strategies.
Beside fitting a statistical model researchers

Statistical methods stand out among these factors because we have a large toolbox for assessing and controlling their contribution to $P(E|¬H)$.
Examples of our ability to exert this control are the setting of type I error rate, the use of corrected fit measures (i.e., adjusted R²), information criteria or cross validation in machine learning.
These are all tools that help us to factor in the tendency of statistical methods to show results even when in fact their are none and are therefore closely related to $P(E|¬H)$  (the probability of observing evidence when in fact there is none).

The point is, the contribution of statistical methods to $P(E|¬H)$ can be formally assesst.
For many statistical models $P(E|¬H)$ can be analytically computed under some assumptions.
For those models or assumptions where this is not possible, one can employ monte carlo simulation to estimate $P(E|¬H)$.
The precision with which researcher can discuss $P(E|¬H)$ has lured the community concerned with research methods into ignoring other factors that are much more uncertain.
We can not hope to resolve this uncertainty, however, we may discuss its implications.

If one chooses to focus only on the contribution of statistical methods to $P(E|¬H)$, one is bound to underestimate it.
Take for example a t test.
Under ideal circumstances (assumption of independence, normality and so forth), it stays true to its type I error rate.
However, researchers might do a number of very reasonable things in the broader context of the study.
They might exclude outliers, choose to drop an item, enlarge their definition of the population to be sampled, translate their questionnaire, impute missing values, or any number of other things, that all carry a small risk that they obtain evidence despite their hypothesis being false.
Even if the t test itself does perfectly maintain its type I error rate these factors have to be added to $P(E|¬H)$.
Statistical properties, therefore, only guarantee a lower bound to the quantity we seek to minimize.
Put simply, the only thing we can conclude with certainty is that $P(E|¬H)$ is not lower than what the statistical model guarantees.

# What to do with uncertainty about the theorethical risk?

@@

Observe, that since $P(E|¬H)$.
